---
title: "Ion concentrations in Bisley, Puerto Rico streams before and after Hurricane Hugo"
format: html
---
## Background
In 1989, Hurricane Hugo hit Puerto Rico, leaving extensive damage in its wake. Long-term monitoring stations in the Luquillo Mountains of Puerto Rico recorded stream water chemisry data in the area before and after the event. This analysis sources and analyzes data from four of those stream sites in Bisley, Puerto Rico to replicate figure 3 in Schaefer et al. ([2000](https://doi.org/10.1017/s0266467400001358)), which explored the effects of hurricane disturbance on these watersheds. In the figure, potassium, nitrate, calcium, magnesium, and ammonia concentration levels in the stream sites are compared before and after the hurricane.

## Data
The raw data were retrieved from the [Environmental Data Initiative](https://doi.org/10.6073/PASTA/F31349BEBDC304F758718F4798D25458). Each csv contains the potassium, nitrate, calcium, magnesium, and ammonia concentrations recorded over time for four sites: 3 sites along the Quebrada Cuenca (referred to as Q1, Q2, Q3), and one at Puente Roto at Rio Mameyes (MPR).

## Methods
The moving average of each ion concentration was calculated for each site in Bisley, Puerto Rico at 9 week intervals. The resulting averages for each site were then plotted from 1988 to 1994 as in the original analysis.

## Results
```{r}
#| include: false
library(tidyverse)
bisley_mov_ave_long <- read_csv("../output/bisley_mov_ave_long.csv")
```
```{r}
#| echo: false
#| warning: false
bisley_mov_ave_long$ion <- factor(
  bisley_mov_ave_long$ion,
  levels = c(
    "k_mgl",
    "no3n_ugl",
    "mg_mgl",
    "ca_mgl",
    "nh4n_ugl"
  )
)
ggplot(
  data = bisley_mov_ave_long,
  mapping = aes(
    x = window_start,
    y = concentration,
    linetype = site_id
  )
) +
  xlim(ymd("1988-01-01"), ymd("1994-06-01")) +
  geom_line() +
  facet_wrap(
    ~ion,
    scales = "free_y",
    ncol = 1,
    strip.position = "left",
    labeller = as_labeller(c(
      ca_mgl = "Ca mg l^-1",
      k_mgl = "K mg l^-1",
      mg_mgl = "Mg mg l^-1",
      nh4n_ugl = "NH4-N ug l^-1",
      no3n_ugl = "NO3-N ug l^-1"
    ))
  ) +
  labs(
    x = "Year",
    y = "Concentration",
    linetype = "Site",
    caption = "Concentrations in Bisley, Puerto Rico streams before and after Hurricane Hugo, 
              9-wk moving averages from 1988 through 1994. (a) potassium, (b) nitrate-N, 
              (c) magnesium, (d) calcium and (e) ammonium-N. The vertical lines mark the time of hurricane disturbance."
  ) +
  geom_vline(
    xintercept = ymd("1989-09-18"),
    linetype = "dashed"
  ) +
  theme_bw()
```
